Cluster Analysis: Identifying Parkinson's Disease Subtypes

### 2015-11-23 writeup/abbreviated/abbreviated.tex
Update writeup with additional pots

\documentclass[letterpaper,12pt]{article}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{numprint}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
% Random packages from
% http://tex.stackexchange.com/questions/50070/landscape-figure-in-latex
% Necessary for sideways pictures
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{epstopdf}
\usepackage{tablefootnote}
% for word wrap verbatim
\usepackage{listings}
\lstset{
   breaklines=true,
   basicstyle=\ttfamily}
% \pagestyle{fancy}
% \lhead{Jesse Mu}
% \rhead{CSCI339 Term Project}
% One more .. since we're in the abbreviated subdir
\graphicspath{ {../../figures/abbreviated/} {../../figures/} }

@@

\subsection{Dataset Description}
951 subjects, 145 metrics, collected 15-4-2012 from Pablo Martinez Mart\'in. Only
19 features used for clustering and/or interpretation.  50 subjects with
missing values of the features to be used in clustering (brought down to 901).
It was decided to not impute the data.

@@

$k$-means clustering with $k = 4$ was tried. Statistics for determining the
optimal number of clusters were used, but were inconclusive: results in
Figure~\ref{tab:numclus}. This probably indicates that the data is not very
well clustered. $k = 2, 3$ provided models that were too simplistic. $k = 5$
did not provide any new information, but rather just fragmented existing
groups.
\begin{table}[h]
  \centering
  \begin{tabular}{l|l}
    Criterion & Optimal $k$ \\
    \hline
    Minimum ASW & 2 \\
    BIC & 18 \\
    SSE Scree Plot & Inconclusive \\
    Gap Statistic & 4, 13? \\
    Affinity Propagation\tablefootnote{$\lambda = 0.98$, q = 0, maxits = 1000,
    convits = 100} & 8 \\

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{kmeans-summaries-4.pdf}
  \caption{Cluster Summaries, $k = 4$}
  \label{fig:kmeans-summaries-4}
\end{figure}

\subsubsection{Interpretation}
$k$-means clustering ($k = 4$) found four clusters. They are numbered (somewhat
confusingly) according to the automatically assigned numbers of the cluster
analysis, and not by any intuitive pattern. With a brief description, they are:

\begin{itemize}
  \item[4.] $(n = 406)$ Mildly affected in all domains.
  \item[1.] $(n = 189)$ Severely affected in nonmotor domains; mildly
    affected in motor domains.
  \item[3.] $(n = 221)$ Severely affected in motor domains; mildly
    affected in nonmotor domains.
  \item[2.] $(n = 88)$ Severely affected in all domains.
\end{itemize}

These cluster results are identical to the four clusters found in van Rooden et
al \cite{vanrooden10}, which was done with a separate dataset using a different
modeling method (expectation-maximization), supporting these subtype
classifications. Unlike van Rooden, mean disease durations differences do exist
between subtypes 4 (mild) and 2 (severe), likely due to further development of
the disease, although the differences between 1 and 3 (nonmotor/motor
predominated) subtypes are insignificant, suggesting different developmental
paths of the disease.

\subsubsection{Statistical Significance Tests, $k = 4$}
Using one-way ANOVA for multiple means, we reject the null
hypothesis that the means are the same with $p < 0.05$ for every variable
\emph{except} pdonset.

Post-hoc analysis using Tukey's HSD to examine statistically significant
differences between individual means is available in Table~\ref{tab:tukeyhsd}.
Only statistically insignificant relations are provided; all other relations
are significant with $p < 0.05$.

\begin{table}[h]
  \centering
  \begin{tabular}{l|l|l}
  Variable & Cluster Relation & $p$ \\
  \hline
  age & 3-1 & 0.724 \\
      & 4-1 & 0.428 \\
  \hline
  sex & 2-1 & 0.849 \\
      & 4-1 & 0.092 \\
      & 3-2 & 0.161 \\
      & 4-2 & 0.827 \\
      & 4-3 & 0.216 \\
  \hline
  pdonset & 2-1 & 0.147 \\
          & 3-1 & 0.370 \\
          & 4-1 & 0.859 \\
          & 3-2 & 0.803 \\
          & 4-2 & 0.305 \\
          & 4-3 & 0.700 \\
  \hline
  durat\_pd & 3-1 & 0.562 \\
  \hline
  cisitot & 3-1 & 0.523 \\
  \hline
  nms\_d1 & 4-3 & 0.333 \\
  \hline
  nms\_d4 & 4-3 & 0.557 \\
  \hline
  nms\_d5 & 4-3 & 0.856 \\
  \hline
  nms\_d8 & 4-3 & 0.122 \\
  \hline
  nms\_d9 & 2-1 & 0.730 \\
         & 4-3 & 0.074 \\
  \hline
  tremor & 4-1 & 0.360 \\
  \end{tabular}
  \caption{Tukey's HSD Insignificant Differences}
  \label{tab:tukeyhsd}
\end{table}

@@

Features ranked by information gain with respect to cluster are available in
Table~\ref{tab:info_gain}. Also, in the 4-cluster decision tree in
Figure~\ref{fig:kmeans-dtree-4}, features are ranked implicitly by importance
in determining clusters. We see, quite naturally, that standard measures of
motor symptoms rank very highly (1, 2, 4, 5) in information gain \emph{except}
termor (12). Similarly, bradykinesia (1) is used as the root node of the
4-cluster decision tree, although other motor symptoms are used further down
the tree, since immediately successive motor symptom decision nodes would, due
to their determination of clusters, be redundant.

@@

Interestingly, demographic information, including durat\_pd, age, sex, and
pdonset, plays almost no role in the determination of these clusters. That the
time of onset of PD or sex nulls important, clinically-relevant questions about
the demographic sources of these different subtypes.

\begin{table}[h]
  \centering
  \begin{tabular}{l|l|l}
    rank & variable & information gain \\
    \hline
    1 & bradykin      & 0.31574672 \\
    2 & rigidity      & 0.29560018 \\
    3 & nms\_d2      & 0.24218407 \\
    4 & cisitot      & 0.22920103 \\
    5 & axial      & 0.22780750 \\
    6 & nms\_d3      & 0.20480570 \\
    7 & nms\_d9      & 0.15782743 \\
    8 & nms\_d7      & 0.15290569 \\
    9 & nms\_d5      & 0.14454931 \\
    10 & nms\_d6      & 0.14025139 \\
    11 & nms\_d1      & 0.13212756 \\
    12 & tremor      & 0.10937168 \\
    13 & nms\_d4      & 0.10710526 \\
    14 & nms\_d8      & 0.10005480 \\
    15 & durat\_pd      & 0.02876190 \\
    16 & age      & 0.02346158 \\
    17 & sex      & 0.00000000 \\
    18 & pdonset      & 0.00000000 \\
  \end{tabular}
  \caption{Features ranked by information gain}
  \label{tab:info_gain}
\end{table}

@@

The interplay between specific symptoms in each of the four clusters was
examined in Figure~\ref{fig:corrplots}, but nothing too interesting was found. There is, perhaps somewhat
interestingly, a somewhat higher correlation between bradykinesia, rigidity,
and nms\_d6 (gastrointestinal), but it was not judged to be significant.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{corrplots.pdf}
  \caption{Correlation plots}
  \label{fig:corrplots}
\end{figure}

\section{Nonmotor-predominant subtype analysis}
\subsection{$k$-means sub-subdivision on Cluster 1}

@@

 The same $k$-determining tests were run on subtype 1 and are displayed
in Table~\ref{tab:numclus-nms}.

\begin{table}[h]
  \centering
  \begin{tabular}{l|l}
    Criterion & Optimal $k$ \\
    \hline
    Minimum ASW & 2 \\
    BIC & 1 (?) \\
    SSE Scree Plot & Inconclusive \\
    Gap Statistic & 3 \\
    Affinity Propagation\tablefootnote{$\lambda = 0.98$, q = 0, maxits = 1000,
    convits = 100} & 5 \\
  \end{tabular}
  \caption{Results of various techniques for determining $k$, applied to
  subtype 1}
  \label{tab:numclus-nms}
\end{table}

Boxplots for $k$-means run for $k = 2, 3, 4$ can
be seen in Figures~\ref{fig:c1-summaries-2},~\ref{fig:c1-summaries-3},
and~\ref{fig:c1-summaries-4}.

@@

An interesting set of subtleties occurs when $k = 2$ and 3. When $k = 2$, the
two groups
are generally somewhat similar, except nms\_d2 and nms\_d3 move in opposite
directions, i.e.\ sub-subtype 2 generally has higher nms\_d2 scores but lower
nms\_d3 scores. In addition, sub-subtype 2 has higher axial scores. This
difference can be once again observed in $k = 3$, where subtype 3 is shown to
have higher nms\_d2 scores, lower nms\_d3 scores, and higher axial scores. This
may indicate a trend in nonmotor-dominated PD for nms\_d2 and nms\_d3 to live
on opposite ends of a spectrum, i.e.\ patients especially severe in one may not
be as severe in the other. This difference can also been seen in
Section~\ref{sub:ova}, when OVA decision trees are discussed for the the
nonmotor-dominated group.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{c1-summaries-2.pdf}
  \caption{Clustering on nonmotor group: $k = 2$}
  \label{fig:c1-summaries-2}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{c1-summaries-3.pdf}
  \caption{Clustering on nonmotor group: $k = 3$}
  \label{fig:c1-summaries-3}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{c1-summaries-4.pdf}
  \caption{Clustering on nonmotor group: $k = 4$}
  \label{fig:c1-summaries-4}
\end{figure}

\section{Modeling}

One further step of this investigation was to produce accurate, practical
models that could be used in a clinical setting to predict the subtype of PD
based on previous clustering results. Cluster assignments were treated as
labels in a supervised classification problem in an attempt to produce useful
models.

@@

While the decision tree in Figure~\ref{fig:kmeans-dtree-4} is useful, it could
be considered overly complicated. Additionally, a model is not necessarily
needed to make simpler diagnoses such as classifying a patient as mildly
affected (subtype 4) or severely affected (subtype 2). One-versus-all (OVA)
decision trees were thus considered, in order to isolate the classification
problem and look at possible distinguishing characteristics of individual
subtypes. These OVA decision trees for all 4 subtypes are located in
Figures~\ref{fig:4va},~\ref{fig:1va},~\ref{fig:3va}, and~\ref{fig:2va}.

\subsubsection{4 (mild)}
This tree is very odd. The root node considers nms\_d9 $< 7.5$ (miscellaneous)
and classifies by far the biggest majority of negative examples when this
symptom is milder, even though the subgroup 4 is the mild subtype. After
intuitively then classifying on the mildness of rigidity the tree then proceeds
to classify once again negative examples based on mildness of nms\_d2 (sleep)
and positive examples solely based on more severe manifestations of nms\_d2,
despite the means of these nonmotor symptoms being quite similar according to
the boxplots.  However, accuracy on the final node is quite poor (61
misclassifications) which could be an indicator that the cluster is not well
defined.

\subsubsection{1 (nonmotor-predominant)}
Quite characteristically, this OVA decision tree clusters first on the severity
of bradykinesia. When mild (i.e.\ $< 2.5$), all of the subsequent decision
nodes ask whether the patient has correspondingly severe forms of primarily
nms\_d3 (mood/cognition), then nms\_d7 (urinary) and nms\_d9 (miscellaneous).

When bradykinesia (and all motor symptoms) are relatively severe, however,
there is still a small subset of patients who classify into subtype 1 by having
severe nms\_d2 (sleep) symptoms. While I am not sure if the sample size is
large enough here (9/34), I propose this could be one of the ``subtypes'' of
the nonmotor-predominant subtype, i.e.\ a group that, when accompanied with
high motor symptoms, generally will have high nonmotor symptoms manifest
primarily in the form of sleep disorders. This aligns with the findings
previously mentioned in Section~\ref{sub:nms-interp} a disjunct between
symptoms nms\_d2 and nms\_d3. As briefly mentioned in that section, axial motor
symptom severity also plays a role. In this tree, mildness of axial symptoms is
the main classifier of 280 negative examples when identifying
nonmotor-predominant
subtypes, indicating it is an important distinguishing factor of this unique
group of 34 nonmotor-predominant PD patients.

More discussion in the
conclusion.

@@

This tree classifies overwhelmingly on severity of bradykinesia, with 476
negative examples when bradykinesia is less than 2.5. The resulting tree is
quite complex, but generally, nodes check again for severity of motor
symptoms (tremor is the next node) and end up classifying positive examples
based on both mildness of nonmotor symptoms and severity of motor symptoms. For
example, in the furthest right branch, once nms\_d2 (as we know, an important
feature) is established to be relatively mild (< 12), the test for subtype 3
involves two more nodes verifying the severity of rigidty and tremor. Similar
behavior can be found in the tree branch testing nms\_d7 and nms\_d6, then
rigidity and axial.

\subsubsection{2 (severe)}

The OVA tree for severely affected in all areas is predictable, testing
entirely on whether or not symptoms (both motor and nonmotor) are relatively
severe. Ppositive nodes always appear to the right (no) of less-than checks.
Interestingly, however, nms\_d4 (percep/hallucinations), previously not of note, is used twice as the
root node of a tree and again further down. As the boxplot display in
Figure~\ref{fig:kmeans-dtree-4} shows, nms\_d4 is perhaps the most
distinguishing symptom of subtype 4 against nonmotor-predominant subtype 1 in
particular, as subtype 1 is relatively mild, in contrast to relatively
comparable levels of severity for other nonmotor symptoms. This shows that
issues with
perception and hallucinations generally occur in only the most severe cases of
PD, and are relatively rare when a patient exhibits a nonmotor-predominant form
of PD.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{dtree-4va-pruned.pdf}
  \caption{Cluster 4 (mild) vs all}
  \label{fig:4va}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{dtree-1va-pruned.pdf}
  \caption{Cluster 1 (nonmotor-dominated) vs all}
  \label{fig:1va}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{dtree-3va-pruned.pdf}
  \caption{Cluster 3 (motor-dominated) vs all}
  \label{fig:3va}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{dtree-2va-pruned.pdf}
  \caption{Cluster 2 (severe) vs all}
  \label{fig:2va}
\end{figure}

\section{Conclusion}

$k$-means clustering on this Parkinsons' Disease data set reveals clusters that
confirm previous computationally-based findings in the field
\cite{vanrooden10}, mainly concerning the identification of four subtypes of
Parkinson's disease: mild, nonmotor-predominant, motor-predominant, and severe.
The most important nonmotor symptoms in determining these clusters were nms\_d2
(sleep) and nms\_d3 (mood/cognition), which echo findings of Fereshtehnejad's
longitudinal study \cite{fereshtehnejad15}. More work

>>>>

\documentclass[10pt]{article}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{numprint}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
% Random packages from
% http://tex.stackexchange.com/questions/50070/landscape-figure-in-latex
% Necessary for sideways pictures
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{epstopdf}
\usepackage{tablefootnote}
% for word wrap verbatim
\usepackage{listings}
\usepackage{caption}
\usepackage{subfig}
\lstset{
   breaklines=true,
   basicstyle=\ttfamily}
% \pagestyle{fancy}
% \lhead{Jesse Mu}
% \rhead{CSCI339 Term Project}
% One more .. since we're in the abbreviated subdir
\graphicspath{ {../../figures/} }
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}

@@

\subsection{Dataset Description}
951 subjects, 145 metrics, collected 15-4-2012 from Pablo Martinez Mart\'in. Only
19 features used for clustering and/or interpretation.  50 subjects with
missing values of the features to be used in clustering (brought down to 901).
It was decided to not impute the data. Data was scaled to $\mu = 0, \sigma =
1$ during clustering and modeling, then unscaled for visualization.

@@

$k$-means clustering with $k = 4$ was tried. Statistics for determining the
optimal number of clusters were used, but were inconclusive: results in
Figure~\ref{tab:numclus}. This probably indicates that the data is not very
well clustered. $k = 2, 3$ provided models that were too simplistic. $k = 5$
did not provide any new information, but rather just fragmented existing
groups.
\begin{table}[h]
  \centering
  \begin{tabular}{l|l}
    Criterion & Optimal $k$ \\
    \hline
    Minimum ASW & 2 \\
    BIC & 18 \\
    SSE Scree Plot & Inconclusive \\
    Gap Statistic & 4 \\
    Affinity Propagation & 8 \\%\tablefootnote{$\lambda = 0.98$, q = 0, maxits = 1000, %convits = 100} & 8 \\

@@

\begin{figure*}[h]
  \centering
  \includegraphics[width=\linewidth]{kmeans-summaries-4.pdf}
  \caption{Cluster Summaries, $k = 4$}
  \label{fig:kmeans-summaries-4}
\end{figure*}

\subsubsection{Interpretation}
$k$-means clustering ($k = 4$) found four clusters. With a brief description,
they are:

% TODO: Figure out where the cluster reordering appears in other graphs
% (kmeans, correlation) and redo those!!!
% AND THE REST OF THE PAPER
% TODO: LEVELS play a big factor - when I factor by reordering levels\dots
\begin{itemize}
  \item[1.] $(n = 406)$ Mildly affected in all domains.
  \item[2.] $(n = 189)$ Severely affected in nonmotor domains; mildly
    affected in motor domains.
  \item[3.] $(n = 221)$ Severely affected in motor domains; mildly
    affected in nonmotor domains.
  \item[4.] $(n = 88)$ Severely affected in all domains.
\end{itemize}

\subsubsection{Statistical Significance Tests, $k = 4$}
For each variable $i$ and cluster means $\mu_i^1, \mu_i^2, \mu_i^3, \mu_i^4$,
we use one-way ANOVA for multiple means and reject the null hypothesis that
$\mu_i^1 = \mu_i^2 = \mu_i^3 = \mu_i^4$ with $p < 0.05$ for every variable
except pdonset.

Post-hoc analysis using Tukey's HSD to examine statistically significant
differences between individual means is available in Table~\ref{tab:tukeyhsd}.
For brevity, only statistically insignificant relations are provided; all other
relations are significant with $p < 0.05$.

\begin{table}[h]
  \centering
  \begin{tabular}{l|l|l}
  Variable & Cluster Relation & $p$ \\
  \hline
age & 2-1 & 0.428 \\
    & 3-2 & 0.724 \\
    \hline
sex & 2-1 & 0.0918 \\
    & 3-1 & 0.216 \\
    & 4-1 & 0.827 \\
    & 4-2 & 0.849 \\
    & 4-3 & 0.161 \\
    \hline
pdonset & 2-1 & 0.859 \\
        & 3-1 & 0.700 \\
        & 4-1 & 0.305 \\
        & 3-2 & 0.370 \\
        & 4-2 & 0.147 \\
        & 4-3 & 0.803 \\
\hline
durat\_pd & 3-2 & 0.562 \\
\hline
cisitot & 3-2 & 0.522 \\
\hline
nms\_d1 & 3-1 & 0.333 \\
\hline
nms\_d4 & 3-1 & 0.557 \\
\hline
nms\_d5 & 3-1 & 0.856 \\
\hline
nms\_d8 & 3-1 & 0.122 \\
\hline
nms\_d9 & 3-1 & 0.0735 \\
       & 4-2 & 0.730 \\
\hline
tremor & 2-1 & 0.360 \\
  \end{tabular}
  \caption{Tukey's HSD Insignificant Differences}
  \label{tab:tukeyhsd}
\end{table}

@@

Features ranked by information gain with respect to cluster are available in
Table~\ref{tab:info_gain}. Also, in the 4-cluster decision tree in
Figure~\ref{fig:kmeans-dtree-4}, features are ranked implicitly by importance
in determining clusters. We see, quite naturally, that standard measures of
motor symptoms rank very highly (ranks 1, 2, 4, 5) in information gain \emph{except}
tremor (12). Similarly, bradykinesia (1) is used as the root node of the
4-cluster decision tree, although other motor symptoms are used further down
the tree, since immediately successive motor symptom decision nodes would, due
to their determination of clusters, be redundant.

@@

Interestingly, demographic information, including durat\_pd, age, sex, and
pdonset, plays almost no role in the determination of these clusters. That the
time of onset of PD or sex is largely irrelevant provides an important negative answer to
clinically-relevant questions about the demographic sources of these different
subtypes.

\begin{table}[h]
  \centering
  \begin{tabular}{r|l|l}
    rank & variable & information gain \\
    \hline
    1 & bradykin      & 0.316 \\
    2 & rigidity      & 0.296 \\
    3 & nms\_d2      & 0.242 \\
    4 & cisitot      & 0.229 \\
    5 & axial      & 0.228 \\
    6 & nms\_d3      & 0.205 \\
    7 & nms\_d9      & 0.158 \\
    8 & nms\_d7      & 0.153 \\
    9 & nms\_d5      & 0.145 \\
    10 & nms\_d6      & 0.140 \\
    11 & nms\_d1      & 0.132 \\
    12 & tremor      & 0.109 \\
    13 & nms\_d4      & 0.107 \\
    14 & nms\_d8      & 0.100 \\
    15 & durat\_pd      & 0.0288 \\
    16 & age      & 0.0235 \\
    17 & sex      & 0.000 \\
    18 & pdonset      & 0.000 \\
  \end{tabular}
  \caption{Features ranked by information gain}
  \label{tab:info_gain}
\end{table}

@@

The interplay between specific symptoms in each of the four clusters was
examined in Figure~\ref{fig:corrplots}. There are two points of note. The first
is that there is a higher correlation in cluster 4 (severe) between overall
severity (cisitot) and bradykinesia and rigidity, illustrated in
Figure~\ref{fig:corrxy-br}.
Second, there exists a somewhat higher correlation between
bradykinesia, rigidity, and nms\_d6 (gastrointestinal) in cluster 4,
illustrated in Figure~\ref{fig:corrxy-d6}. These differences are statistically
significant; correlation tests are located in Table~\ref{tab:cortests}. I am
unsure of the significance or proper interpretation of these results.

\begin{table*}[h]
  \centering
  \begin{tabular}{l|l|l|l}
 Cluster & Variables & 95\% CI & $p$ \\
    \hline
1 & bradykin, cisitot&[-0.0225, 0.171]&0.131 \\
  & rigidity, cisitot&[-0.000406, 0.192]&0.0510 \\
  & bradykin, nms\_d6&[-0.0634, 0.131]&0.493 \\
  & rigidity, nms\_d6&[-0.101, 0.0932]&0.934 \\
  \hline
2 & bradykin, cisitot&[0.0786, 0.351]&$0.00248^{(**)}$ \\
  & rigidity, cisitot&[-0.215, 0.069]&0.310 \\
  & bradykin, nms\_d6&[-0.152, 0.133]&0.897 \\
  & rigidity, nms\_d6&[-0.123, 0.163]&0.781 \\
  \hline
3 & bradykin, cisitot&[0.0995, 0.350]&$0.000620^{(***)}$ \\
  & rigidity, cisitot&[0.0687, 0.322]&$0.00298^{(**)}$ \\
  & bradykin, nms\_d6&[0.0350, 0.292]&$0.0134^{(*)}$ \\
  & rigidity, nms\_d6&[-0.0846, 0.179]&0.478 \\
  \hline
  4 & bradykin, cisitot&[0.454, 0.724]&${3.97\e{-10}}^{(***)}$ \\
& rigidity, cisitot&[0.375, 0.675]&${4.99\e{-08}}^{(***)}$ \\
  & bradykin, nms\_d6&[0.297, 0.624]&${2.60\e{-06}}^{(***)}$ \\
  & rigidity, nms\_d6&[0.278, 0.611]&${6.43\e{-06}}^{(***)}$ \\
  \end{tabular}
  \caption{Correlation tests. (*) $p < 0.05$, (**) $p < 0.01$, (***), $p < 0.001$}
  \label{tab:cortests}
\end{table*}

\begin{figure}[h]
  \centering
  \subfloat[][]{%
  \includegraphics[clip,width=0.8\linewidth]{bradykin-v-cisitot.pdf}
  \label{bradyv}
}

\subfloat[][]{%
  \includegraphics[clip,width=0.8\linewidth]{rigidity-v-cisitot.pdf}
  \label{rigidv}
}
  \caption{Relationship between \protect\subref{bradyv} bradykinesia,
  \protect\subref{rigidv} rigidity and overall severity (cisitot). Shaded band
is 95\% confidence interval.}
  \label{fig:corrxy-br}
\end{figure}

\begin{figure}[h]
  \centering
  \subfloat[][]{%
  \includegraphics[clip,width=0.8\linewidth]{bradykin-v-nms_d6.pdf}
  \label{brady6}
}

\subfloat[][]{%
  \includegraphics[clip,width=0.8\linewidth]{rigidity-v-nms_d6.pdf}
  \label{rigid6}
}
  \caption{Relationship between \protect\subref{brady6} bradykinesia,
  \protect\subref{rigid6} rigidity and nms\_d6 (gastrointestinal). Shaded band
is 95\% confidence interval.}
  \label{fig:corrxy-d6}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.68\linewidth]{corrplots.pdf}
  \caption{Correlation plots}
  \label{fig:corrplots}
\end{figure}

\section{Nonmotor-predominant subtype analysis}
\subsection{$k$-means sub-subdivision on Cluster 2}
% TODO: Synthesize changed subtype 2 analysis with discussion HERE and in
% conclusion. Make sure to reread and ensure that all plots have been updated
% (in particular, the 4 cluster decision tree?). Incorporate some stuff from
% standard progression of motor diseases paper and the new nms subtype paper.

@@

The same $k$-determining tests were run on subtype 2 and are displayed
in Table~\ref{tab:numclus-nms}.

\begin{table}[h]
  \centering
  \begin{tabular}{l|l}
    Criterion & Optimal $k$ \\
    \hline
    Minimum ASW & 2 \\
    BIC & 1 (?) \\
    SSE Scree Plot & Inconclusive \\
    Gap Statistic & 3 \\
    Affinity Propagation\tablefootnote{$\lambda = 0.98$, q = 0, maxits = 1000,
    convits = 100} & 5 \\
  \end{tabular}
  \caption{Results of various techniques for determining $k$, applied to
  subtype 2}
  \label{tab:numclus-nms}
\end{table}

Boxplots for $k$-means run for $k = 2, 3, 4$ can
be seen in Figures~\ref{fig:c2-summaries-2},~\ref{fig:c2-summaries-3},
and~\ref{fig:c2-summaries-4}. Clusters are ordered by increasing cisitot.

@@

An interesting set of subtleties occurs when $k = 2$ and 3. When $k = 2$, the
two groups are divided generally by PD severity (see cisitot and especially
axial). The specific symptoms of the two groups follow this trend, except
nms\_d3 and tremor, which are actually decreasing, and other symptoms like
rigidity, nms\_d4, and nms\_d9, which are more indeterminate.

When $k = 3$, the symptoms that continue show a non-monotonically increasing
trend are nms\_d2, tremor, and rigidity scores, where patients in the 3rd
subtype exhibit lower severities. nms\_d4 and nms\_d9 differences turn out to be
not as pronounced.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{c2-summaries-2.pdf}
  \caption{Clustering on nonmotor group: $k = 2$}
  \label{fig:c2-summaries-2}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{c2-summaries-3.pdf}
  \caption{Clustering on nonmotor group: $k = 3$}
  \label{fig:c2-summaries-3}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{c2-summaries-4.pdf}
  \caption{Clustering on nonmotor group: $k = 4$}
  \label{fig:c2-summaries-4}
\end{figure}

\section{Further modeling}

One further step of this investigation was to produce accurate, practical
models that could be used in a clinical setting to predict the subtype of PD
based on previous clustering results. Cluster assignments obtained from
previous $k$-means investigation were treated as labels in a supervised
classification problem in an attempt to produce useful and easily interpretable
models.

@@

While the decision tree in Figure~\ref{fig:kmeans-dtree-4} is useful, it could
be considered overly complicated. Additionally, a model is not necessarily
needed to make simpler diagnoses such as classifying a patient as mildly
affected (subtype 1) or severely affected (subtype 4). One-versus-all (OVA)
decision trees were thus considered, in order to isolate the classification
problem and look at possible distinguishing characteristics of individual
subtypes. These OVA decision trees for all 4 subtypes are located in
Figures~\ref{fig:1va},~\ref{fig:2va},~\ref{fig:3va}, and~\ref{fig:4va}. Trees
are pruned by selecting the version of tree with the minimum 10-fold
cross-validated error.

\subsubsection{1 (mild)}
The tree for the mild subtype classifies mainly based on negative responses to
nodes asking whether the patient has a relatively severe manifestation of a
symptom. The majority of examples are classified by following the bradykinesia
$<$ 2.5, which subsequently tests the severity of several nonmotor symptoms.
Most of subtype 1 patients that score relatively mildly on these scales are
classified this way. There are also small populations of patients who 1) score
% http://www.ncbi.nlm.nih.gov/pubmed/22233888
higher on bradykinesia but lower with axial, tremor, rigidity, and nms\_d2 and
2) score higher in nms\_d2 (sleep) but lower with nms\_d7 (urinary).
% This tree is very odd. The root node considers nms\_d9 $< 7.5$  and nms\_d3 (miscellaneous)
% and classifies by far the biggest majority of negative examples when this
% symptom is milder, even though the subgroup 1 is the mild subtype. After
% intuitively then classifying on the mildness of rigidity (rigidity >= 3.5) the tree then proceeds
% to classify once again negative examples mostly based on mildness of nms\_d2
% (sleep)
% and positive examples based on more severe manifestations of nms\_d2,
% despite the means of these nonmotor symptoms being quite similar according to
% the boxplots.  However, accuracy on the final nodes are quite poor, which could
% be an indicator that the cluster is not well defined.

\subsubsection{2 (nonmotor-predominant)}
The decision tree for the nonmotor-predominant subtype is quite simple.
Interestingly, although nms\_d9 (miscellaneous) is not the most important
nonmotor symptom, since the information gain is less than nms\_d2 and nms\_d3
and it does not appear very high in the 4-class decision tree, it is used as
the root node of this decision tree, classifying over half of the negative
examples based on whether the subject has a low severity of miscellaneous
symptoms (nms\_d9 $<$ 7.5)\footnote{Recall that 0 $<$ nms\_d9 $<$ 48.}. This could be an indication that nonmotor-predominant PD patients do
indeed have a wide manifestation and variety of nonmotor symptoms. After
classifying on nms\_d9, the tree then classifies negatiave examples as having
rigidity $\geq$ 3.5, an example of how subtype 2 patients have relatively low motor
symptoms. Finally, the tree classifies on the nonmotor symptom with the most
information gain, nms\_d2, where patients $\geq$ 7.5 are classified as falling into
subtype 2.

% Quite characteristically, this OVA decision tree clusters first on the severity
% of bradykinesia. When mild (i.e.\ bradykin $< 2.5$), all of the subsequent decision
% nodes ask whether the patient has correspondingly severe forms of primarily
% nms\_d3 (mood/cognition), then nms\_d7 (urinary) and nms\_d9 (miscellaneous).

% When bradykinesia (and all motor symptoms) are relatively severe, however,
% there is still a small subset of patients who classify into subtype 2 by having
% severe nms\_d2 (sleep) symptoms. While I am not sure if the sample size is
% large enough here (9/34), I propose this could be one of the ``subtypes'' of
% the nonmotor-predominant subtype, i.e.\ a group that, when accompanied with
% high motor symptoms, generally will have high nonmotor symptoms manifest
% primarily in the form of sleep disorders. This aligns with the findings
% previously mentioned in Section~\ref{sub:nms-interp} a disjunct between
% symptoms nms\_d2 and nms\_d3. As briefly mentioned in that section, axial motor
% symptom severity also plays a role. In this tree, mildness of axial symptoms is
% the main classifier of 280 negative examples when identifying
% nonmotor-predominant
% subtypes, indicating it is an important distinguishing factor of this unique
% group of 34 nonmotor-predominant PD patients.

% More discussion on this decision tree, synthesized with the other results of
% this research, is in the conclusion.

@@

This tree classifies overwhelmingly on severity of bradykinesia, with 476
negative examples when bradykinesia is less than 2.5. The resulting tree is
quite complex, but generally, nodes check again for severity of motor
symptoms (tremor is the next node) and end up classifying positive examples
based on both mildness of nonmotor symptoms and severity of motor symptoms. For
example, in the furthest right branch, once nms\_d2 (as we know, an important
feature) is established to be relatively mild ($<$ 12), the test for subtype 3
involves several more nodes verifying the severity of rigidty, tremor, and
axial, and the mildness of nms\_d7 (urinary).

\subsubsection{4 (severe)}

The OVA tree for patients severely affected in all areas is predictable, testing
entirely on whether or not symptoms (both motor and nonmotor) are relatively
severe. Positive nodes always appear to the right (no) of less-than checks.
Interestingly, however, nms\_d4 (percep/hallucinations), previously not of note, is used twice as the
root node of a tree and again further down. As the boxplot display in
Figure~\ref{fig:kmeans-summaries-4} shows, nms\_d4 is perhaps the most
distinguishing symptom of subtype 4 against nonmotor-predominant
subtype 2 in particular, as subtype 2 has relatively mild percep/hallucination
symptoms, in contrast to the comparable levels of severity for other nonmotor
symptoms in both groups. This shows that issues with perception and
hallucinations generally occur in only the most severe cases of PD, and are
relatively rare when a patient exhibits a nonmotor-predominant form of PD.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\linewidth]{dtree-1va-pruned.pdf}
  \caption{Cluster 1 (mild) vs all}
  \label{fig:1va}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\linewidth]{dtree-2va-pruned.pdf}
  \caption{Cluster 2 (nonmotor-dominated) vs all}
  \label{fig:2va}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\linewidth]{dtree-3va-pruned.pdf}
  \caption{Cluster 3 (motor-dominated) vs all}
  \label{fig:3va}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\linewidth]{dtree-4va-pruned.pdf}
  \caption{Cluster 4 (severe) vs all}
  \label{fig:4va}
\end{figure}

\subsection{Different angles of exploration: 2 and 4, 2 and 3 vs rest}

There are many more interesting questions to be asked when examining the
relationship between these clusters. One thing that may be helpful in
understanding the relationship between the clusters is exploring different
groupings of clusters for decision trees. The trees in
Figure~\ref{fig:dtree-2and4va-pruned} and~\ref{fig:dtree-2and3va-pruned} are
preliminary examples of this kind
of exploration.
